Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-06 paper/paper.tex
Comments abstract.

\begin{abstract}
The default mode network (DMN) is believed to subserve the human
baseline mental activity.
%
Many research streams converge to its likely role in evolutionarily adaptive
envisioning of mental scenarios to simulate the future.
The DMN may be dedicated to continuous autobiographical memory retrieval,
generation of hypothetical outcomes,
and reward contingency evaluation when letting the mind go.
Such functional roles offer explanations of
its highest energy consumption in the brain and
its intimate coupling with conscious awareness.
%
This concept paper proposes a process model that describes
\textit{how} the DMN may actually implement continuous
assessment and prediction of the environment to guide action choices.
DMN function is recast in the mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
We argue that our formal account of DMN function
naturally accommodates previously proposed cognitive accounts as special cases and
parsimoneously explains various experimental findings in animals and humans.
%
Formal models of the neural computations realized in the DMN
could offer access routes to the statistical regularities
of human predictive behavior.
\end{abstract}

@@

\begin{mdframed}
  \textbf{Time-invariance assumption:} It is assumed that the underlying MDP is stationary --i.e has reached steady-state conditions--
  so that the dynamics of the process are equivalent over sufficiently long time windows of equal length.
\end{mdframed}

@@

\subsection{The Q-values, state-values, and the advantage function}
We can now formulate the purpose of the DMN
as finding a policy $\pi$ for the human agent that maximizes the
expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a].
\end{equation}
In other words,
the Q-value $Q^{\pi}(s,a)$ corresponds to the expected reward
over all possible action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$,
and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utility of taking action $a$ in this situation ?".
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior,
potentially driven by the DMN in human agents.

One can also define the state-value function

\begin{eqnarray}
  \V^\pi(s) := \mathbb E_{a \sim \pi}Q^\pi(s,a) = \mathbb E [G^\pi|s_0=s],
\end{eqnarray}
which measures the value of each state $s \in \mathcal S$ under the policy $\pi$. This serves
as a baseline value for any action. Finally, the advantage function is defined by
\begin{equation}
A^\pi(s,a) :=  Q^\pi(s,a) - V^\pi(s).
\end{equation}

% The goal of the DMN is then
% to find then to produce a policy $\pi$ which maximizes the expected cumulative reward $Q^\pi(s_0,\pi(a_0))$
% given an initial state $s_0$.

@@

\subsubsection{Q-learning} Q-learning (Watkins \& Dayan, 1992) generates a
\textit{greedy policy}
\textbf{XXX please add 1-2 sentences why this is interesting in the current
context. This statement appears a little alone as it stands.}
\begin{equation}
  \pi^*(s) = \argmax_{a \in \mathcal A}{Q}^*(s, a),
  \label{eq:qlearning}
\end{equation}
which maps each state $s$ to the optimal action which should be taken at that state.
Here $Q^*(s, a) := \max_{\pi}Q^\pi(s,a)$, the optimal Q-value over all policies.
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The \textit{Bellman equation} \citep{sutton1998reinforcement} then takes the simple form
\begin{equation}
  \begin{split}
    Q^*(s_0, a_0) &= \mathbb E_{s_1 \sim p(s_1|s_0,a)} [r(s,a) + \gamma \max_{a_1 \in \mathcal A}Q^*(s_1, a_1)]
    \\
    &= \mathbb E_{s_1 \sim p(s_1|s_0,a_0)} [r(s_0,a_0) + \gamma Q^*(s_1, \pi(s_1))].
  \end{split}
  \label{eq:bellman}
\end{equation}
This consistency equation for the optimal (over all policies) Q-value function provides
a recursive decomposition of optimal behavior by dividing the Q-value function into the immediate
reward and the discounted rewards of the upcoming states. The optimal Q-value operator $Q^*$
is a fixed point for this equation. A complicated dynamic programming optimization can be
broken into simpler sub-problems at different time points,
which has often been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in \eqref{eq:bellman} the sampling is done only over things which
depend on the environment, and so can be learned off-policy by observing state transitions
triggered by another behavioral policy, which can be stochastic.

>>>>

\begin{abstract}
The default mode network (DMN) is believed to subserve the human
baseline mental activity.
%
Many research streams converge to its likely role in evolutionarily adaptive
envisioning of mental scenarios to simulate the future.
\suggestremove{The DMN may be dedicated to continuous autobiographical memory retrieval,
generation of hypothetical outcomes,
and reward contingency evaluation when letting the mind go.}
Such functional roles offer explanations of
its highest energy consumption in the brain and
its intimate coupling with conscious awareness
\textbf{(XXX Guillaume: It escaladed quickly.)}.
%
This concept paper proposes a process model \textbf{(XXX Guillaume: It's not clear what a "process model" is for most people?)} that describes
\textit{how} the DMN may actually implement continuous
assessment and prediction of the environment to guide action choices.
DMN function is recast in the mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
We argue that our formal account of DMN function
naturally accommodates \suggestremove{previously} proposed cognitive accounts as special cases and
parsimoneously explains \suggestremove{various} experimental findings in \suggestadd{both} animals and humans.
%
Formal models of the neural computations realized in the DMN
could offer access routes to the statistical regularities
of human predictive behavior \textbf{(XXX Guillaume: Maybe too wordy\ldots)}.
\end{abstract}

@@

\begin{mdframed}
  \textbf{Time-invariance assumption:} It is assumed that the MDP has reached steady-state conditions so that
  the dynamics of the process are equivalent over sufficiently long time windows of equal length.
\end{mdframed}

@@

\subsection{The Q-value: A goodness measure for action choices}
We can now formulate the purpose of the DMN
as finding a policy $\pi$ for the human agent that maximizes the
expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a,a_1 \sim \pi,a_2 \sim \pi,\ldots].
\end{equation}
In other words,
the Q-value $Q^{\pi}(s,a)$ corresponds to the expected reward
over all possible action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$,
and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utility of taking action $a$ in this situation ?".
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior,
potentially driven by the DMN in human agents.
% The goal of the DMN is then
% to find then to produce a policy $\pi$ which maximizes the expected cumulative reward $Q^\pi(s_0,\pi(a_0))$
% given an initial state $s_0$.

@@

\subsubsection{Q-learning} Q-learning (Watkins \& Dayan, 1992) generates a
\textit{greedy policy}
\textbf{XXX please add 1-2 sentences why this is interesting in the current
context. This statement appears a little alone as it stands.}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The \textit{Bellman equation} \citep{sutton1998reinforcement} then takes the simple form
\begin{equation}
  \begin{split}
    Q^\pi(s, a) &= \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q^\pi(s', a')]
    \\
    &= \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))].
  \end{split}
  \label{eq:bellman}
\end{equation}
This \textit{fixed-point} equation provides a recursive decomposition of optimal behavior by dividing the value function \textbf{XXX please more explicitly introduce
the notion of value function}
into the immediate reward
and the discounted rewards of the upcoming states.
A complicated dynamic programming optimization can be
broken into simpler sub-problems at different time points,
which has often been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in \eqref{eq:bellman} the sampling is done only over things which
depend on the environment, and so can be learned off-policy by observing state transitions
triggered by another behavioral policy, which can be stochastic.
